The Chemistry Development Kit (CDK). 3. Atom typing, Rendering, Molecular Formula, and Substructure Searching

### 2015-12-23 article.tex
added signature fingerprints reference to fp table
Patches from Friday evening

\begin{abstract}
% Do not use inserted blank lines (ie \\) until main body of text.
\paragraph*{Background:}
Cheminformatics is a well-established field with many applications in chemistry,
biology, drug discovery, and others. The Chemistry Development Kit (CDK) has
become a widely used Open Source cheminformatics toolkit, providing various models to represent
chemical structures, of which the chemical graph is essential. However,
in the first five years of the project increased so much in size that
interdependencies between components grew unmanageable large, resulting
in unpredictable instabilities.
\paragraph*{Results:}
We here report the improvements of the CDK since the 1.2 release series made
to accomodate both the increased complexity of the library, as well as
siginificant improvements of and additions to the functionality of the
library. We here outline how the CDK evolved since the previous report and
discuss the quality control approached we have adopted to ensure stability.
Additionally, new APIs have been introduced for atom type perception,
substructure searching, molecular fingerprints, rendering of molecules, and
handling of molecular formulas.
\paragraph*{Conclusions:}
With this paper we have shown the continued effort to provide a free, Open Source
cheminformatics library, and show that such colleborative projects can survive for
a long period. We have taken advantage from the community support, and show
that an open source cheminformatics project can act as a peer reviewed
publishing platform for scientific computing software.
\end{abstract}

@@

The chemical formula is the basic/simple chemical representation of a compound. It defines the number of atoms/isotopes and type of elements that constitute a chemical compound without describing how atoms are bonded.
It is also the first step toward the identification of the chemical structure from a unidentified compound.

@@

  \subsubsection*{Partial charges and Delocalization}

Partial atomic charges is an electronic property of atoms defining the asymmetric distribution of electrons in chemical bonds.
It is used as a quantitative correlation with certain compound's physical and chemical properties.

CDK can
\begin{itemize}
\item generate all resonance structures from a given compound 
\item calculate the topological and electronic weight value of each of the resonance structures
\item calculate the partial atomic charge from a given structure 
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
  \ifthenelse{\boolean{publ}}{\small}{}
Niels Out and Stefan Kuhn are acknowledged for their contributions to the project.

@@

    \begin{minipage}{1\textwidth}
    \renewcommand*{\thempfootnote}{\fnsymbol{mpfootnote}}
    \centering
    \rowcolors{1}{}{lightgray}
    \begin{tabular}{lcclcc}
    \toprule
                             & Bit version  & Count version & CDK version & Default Size    & References \\
  CirkularFingerprinter      & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $1\,024$ / $2^{32}$%
\footnote[1]{For the CirkularFingerprinter the bit version is folded to 1024 whereas the count version is unfolded}
                                                                                            & \cite{rogers2010extended, Clark2014} \\
  EStateFingerprinter        & $\checkmark$ &               & 1.2.0       & $79$            & \cite{Hall1995} \\
  ExtendedFingerprinter      & $\checkmark$ &               & 1.0         & $1\,024$        & \\
  Fingerprinter              & $\checkmark$ &               & 1.0         & $1\,024$        & \\
  GraphOnlyFingerprinter     & $\checkmark$ &               & 1.0         & $1\,024$        & \\
  HybridizationFingerprinter & $\checkmark$ &               & 1.4.0       & $1\,024$        & \\
  KlekotaRothFingerprinter   & $\checkmark$ &               & 1.4.6       & $4\,860$        & \cite{Klekota2008} \\
  LingoFingerprinter         & $\checkmark$ &               & 1.6.0 ?     & NA%
\footnote[2]{The Lingofingerprinter does not have a default size}               
                                                                                            & \cite{vidal2005lingo} \\
  MACCSFingerprinter         & $\checkmark$ &               & 1.2.0       & $166$           & \\
  PubchemFingerprinter       & $\checkmark$ &               & 1.4.0       & $881$           & \cite{pubchemFP} \\
  ShortestPathFingerprinter  & $\checkmark$ &               & 1.6.0 ?     & $1\,024$        & \\
  SignatureFingerprinter     & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $2^{32}$        & [pending] \\
  SubstructureFingerprinter  & $\checkmark$ &               & 1.0         & $307$           & \\
  \bottomrule
    \end{tabular}
    \end{minipage}

>>>>

\begin{abstract}
% Do not use inserted blank lines (ie \\) until main body of text.
\paragraph*{Background:}
Cheminformatics is a well-established field with many applications in chemistry,
biology, drug discovery, and others. The Chemistry Development Kit (CDK) has
become a widely used Open Source cheminformatics toolkit, providing various models to represent
chemical structures, of which the chemical graph is essential. However,
in the first five years of the project increased so much in size that
interdependencies between components grew unmanageable large, resulting
in unpredictable instabilities.
\paragraph*{Results:}
We here report improvements to the CDK since the 1.2 release series made
to accomodate both the increased complexity of the library, as well as
siginificant improvements of and additions to the functionality of the
library. Second, we outline how the CDK evolved with respect to
quality control and the approach we have adopted to ensure stability, including
a peer review mechanism.
Additionally, a selection of the new APIs that have been introduced will
be discussed: atom type perception, substructure searching, molecular
fingerprints, rendering of molecules, and handling of molecular formulas.
\paragraph*{Conclusions:}
With this paper we have shown the continued effort to provide a free, Open Source
cheminformatics library, and show that such colleborative projects can exist over
a long period. We have taken advantage from the community support, and show
that an open source cheminformatics project can act as a peer reviewed
publishing platform for scientific computing software.
\end{abstract}

@@

The chemical formula is the basic/simple chemical representation of a compound.
It defines the number of isotopes or elements that constitute a the composition
of a compound without describing how atoms are bonded. With the rise of metabolomics,
it has become increasingly relevant to have full support in cheminformatics
libraries~\cite{RojasChertoXXX}.
It is also the first step
toward the identification of the chemical structure from a unidentified compound.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
\ifthenelse{\boolean{publ}}{\small}{}
The authors acknoledge the great number of people who have contributed smaller
and larger contributions to the CDK library. A full list of contributors is
found in the AUTHORS file~\cite{AUTHORS}.

@@

    \begin{minipage}{1\textwidth}
    \renewcommand*{\thempfootnote}{\fnsymbol{mpfootnote}}
    \centering
    \rowcolors{1}{}{lightgray}
    \begin{tabular}{lcclcc}
    \toprule
                             & Bit version  & Count version & CDK version & Default Size    & References \\
  CirkularFingerprinter      & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $1\,024$ / $2^{32}$%
\footnote[1]{For the CirkularFingerprinter the bit version is folded to 1024 whereas the count version is unfolded}
                                                                                            & \cite{rogers2010extended, Clark2014} \\
  EStateFingerprinter        & $\checkmark$ &               & 1.2.0       & $79$            & \cite{Hall1995} \\
  ExtendedFingerprinter      & $\checkmark$ &               & 1.0         & $1\,024$        & \\
  Fingerprinter              & $\checkmark$ &               & 1.0         & $1\,024$        & \\
  GraphOnlyFingerprinter     & $\checkmark$ &               & 1.0         & $1\,024$        & \\
  HybridizationFingerprinter & $\checkmark$ &               & 1.4.0       & $1\,024$        & \\
  KlekotaRothFingerprinter   & $\checkmark$ &               & 1.4.6       & $4\,860$        & \cite{Klekota2008} \\
  LingoFingerprinter         & $\checkmark$ &               & 1.6.0 ?     & NA%
\footnote[2]{The Lingofingerprinter does not have a default size}               
                                                                                            & \cite{vidal2005lingo} \\
  MACCSFingerprinter         & $\checkmark$ &               & 1.2.0       & $166$           & \\
  PubchemFingerprinter       & $\checkmark$ &               & 1.4.0       & $881$           & \cite{pubchemFP} \\
  ShortestPathFingerprinter  & $\checkmark$ &               & 1.6.0 ?     & $1\,024$        & \\
  SignatureFingerprinter     & $\checkmark$ & $\checkmark$  & 1.6.0 ?     & $2^{32}$        & \cite{signaturefingerprints} \\
  SubstructureFingerprinter  & $\checkmark$ &               & 1.0         & $307$           & \\
  \bottomrule
    \end{tabular}
    \end{minipage}
